Think deep for reef resilience

### 2025-08-09 paper/paper_draft.qmd
Updated number of reefs and citations
@ConnectedSystems I have made a couple of new comments about depth numbers

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority [@greatba2022]. An
indicative map of the GBR marine park including the Marine Park regions and which bioregion
each reef belongs to is shown in @fig-region-map. In total, the dataset represents 3806 reef
areas across the GBR marine park. Reefs that are outside the GBRMPA management areas or are
not associated with a GBRMPA reefal bioregion **\[TI: Ben, cite the bioregion data
source\]** were removed from the investigation, leaving 3531 reefs. Annually and spatially
averaged Degree Heating Week (DHW) data were used to represent future thermal stress at each
reef location. Code used to calculate DHW data and perform analyses involved in this
investigation can be found at **insert repo citation**.

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points). The green polygon represents the Far Northern management area,
the yellow polygon represents the Cairns/Cooktown management area, the pink polygon
represents the Townsville/Whitsunday management area and the blue polygon represents the
Mackay/Capricorn management area. The centroid of each reef is represented as a point for
visual clarity, and is coloured according to its GBRMPA reefal bioregion. An annotation
represents the approximate location of the Australian Institute of Marine Science Townsville
site at Cape
Cleveland.](../outputs/ADRIA_results/HighResCoralStress/figs/region_map.png){#fig-region-map}

@@

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled
1km^2^ resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4 and are a part of the Sixth Phase of the
Coupled Model Intercomparison Project (CMIP6) based on evaluation conducted by Grose et al.
[-@grose2023]. These GCMs were found to rank highly in the recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. [-@grose2023] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region. An overview of the selected GCMs are provided in @tbl-gcms. The process
described herein was conducted for all reefs within the GBR Marine Park to create DHW
projections spanning from 2025 to 2075.

@@

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of positive HotSpot SST Anomaly values, expressed in units of °C-weeks. The
12-week accumulation period reflects the cumulative nature of coral thermal stress. Brief
temperature spikes may be tolerable, but sustained or repeated exposure to elevated
temperatures causes progressively greater physiological stress and bleaching severity. DHW
values integrate both the magnitude and duration of thermal stress: a moderate anomaly
(e.g., 1°C) sustained for 12 weeks produces the same DHW value as a severe anomaly (e.g.,
3°C) lasting 4 weeks, though the biological impacts may differ. DHW estimates for each
indicative reef (see @greatba2022) were then determined by taking the mean DHW value for all
pixels touching each reef outline.

::: {#tbl-gcms}
| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Results found in |
|------------------------|----------------------|--------------------------|----------------------|
| EC-Earth3-Veg | 4.31 | High | **insert repo citation** |
| ACCESS-CM2 | 4.72 | High | **insert repo citation** |
| ACCESS-ESM1-5 | 3.87 | Intermediate | Main text |
| GFDL-CM4 | 2.9 | Low | **insert repo citation** |
| NorESM2-MM | 2.5 | Low | **insert repo citation** |

@@

Reef depth characteristics were extracted using 10m resolution bathymetry data from the
Great Barrier Reef Marine Park Authority (GBRMPA [-@gbr10gb2020]). For each reef, bathymetry
data (in raster format) were spatially matched with reef polygon features through geometric
intersection analysis. Depth values within each reef polygon were summarized using zonal
statistics to calculate minimum, mean, median, maximum, and standard deviation of depth
values. All depth values were converted to positive values (depths below sea level). The
median depth was used to provide an indicative depth value for each reef.

As bathymetry data are derived from satellite imagery they can be limited in their alignment
with the established GBRMPA reef outlines used in the investigation to characterise reef
locations. Quality control flags were assigned to identify potential data limitations: reefs
with no overlapping bathymetry data, reefs with minimum depth values above sea level
(indicating potential data artifacts), and reefs where bathymetry pixels covered less than
5% of the reef polygon area. These reefs with insufficient data coverage were removed from
analyses (842 reefs of 3806 reefs, or 22%) **\[TI: BG to update\]**.

While data adopted for the given assessment are said to be accurate up to 15m from Mean Sea
Level [@gbr10gb2020], there are many sources of uncertainty in the depth estimates used for
this study. There is potential misalignment between GBRMPA reef outlines, as well as the
usual challenges of deriving data from satellite imagery. In the case of bathymetry,
estimates are subject to different bottom types (rock, sand, seagrass, etc) and how these
affect reflectance, the water penetration depth of satellite sensors, influence of local
conditions (e.g., turbidity and wave activity), and many other factors. The resulting depth
estimates are therefore highly uncertain themselves and are only adopted as a useful
indication rather than an accurate and precise report of reef depths across the GBR.

@@

Current reef restoration and intervention-investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable restoration
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One of the main
constraints on deployment depth is visual identification of target sites from the surface
for deployment and ongoing monitoring, in combination with factors such as deployment method
and tourism return value. Our results suggest that of the reefs analysed within the 1-10m
depth range (1800 reefs), 74% (1328) of reefs fall into the low and medium clusters (across
all GCMs at the GBRMPA management region scale), with many unable to maintain positive
carbonate budgets under future conditions. Reef intervention planning is a complex and
challenging decision making process [@anthony2020], and our results suggest that deployment
site depth is an important factor to consider when assessing the long-term resilience of
deployment sites.

>>>>

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority (GBRMPA)
[@GBRMPA2022]. An indicative map of the GBR marine park including the Marine Park
management regions and which reefal bioregion each reef belongs to is shown in
@fig-region-map [@GBRMPA2024; @GBRMPA2001]. This investigation adopts the GBR reef definition
used in Bozec et al. [-@bozec2025], which includes 3806 reefs across the GBR. Reefs that are not
associated with GBRMPA management areas or reefal bioregions were removed from the
investigation. Additionally, reefs with poor quality bathymetry data coverage or fewer than
three reefs per bioregion were removed from analyses (see Methods - Depth estimation). As a
result, a total of 2429 individual reefs were included in this investigation. Annually and
spatially averaged Degree Heating Week (DHW) data were used to represent future thermal
stress at each reef location. Code used to calculate DHW data and perform analyses involved
in this investigation can be found at **insert repo citation**.

\[**BG: Updated number of reefs and introduced original set of reefs with 3806 to set scene for justified removal of reefs.**\]

\[**BG: I am having issues with citations not appearing for datasets like GBRMPA ones.**\]

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points). The green polygon represents the Far Northern management area,
the yellow polygon represents the Cairns/Cooktown management area, the pink polygon
represents the Townsville/Whitsunday management area and the blue polygon represents the
Mackay/Capricorn management area [@GBRMPA2024]. The centroid of each reef is represented as a
point for visual clarity, and is coloured according to its GBRMPA reefal bioregion
[@GBRMPA2001]. All 2429 reefs included in this investigation are represented in this figure. An
annotation represents the approximate location of the Australian Institute of Marine Science
Townsville site at Cape Cleveland.](../outputs/ADRIA_results/HighResCoralStress/figs/region_map.png){#fig-region-map}

@@

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled
1km^2^ resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4 and are a part of the Sixth Phase of the
Coupled Model Intercomparison Project (CMIP6) based on evaluation conducted by Grose et al.
[-@grose2023]. These GCMs were found to rank highly in the recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. [-@grose2023] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region. An overview of the selected GCMs are provided in @tbl-gcms. The process
described herein was conducted for all 2429 reefs to create DHW projections spanning from
2025 to 2075.

@@

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of positive HotSpot SST Anomaly values, expressed in units of °C-weeks. The
12-week accumulation period reflects the cumulative nature of coral thermal stress. Brief
temperature spikes may be tolerable, but sustained or repeated exposure to elevated
temperatures causes progressively greater physiological stress and bleaching severity. DHW
values integrate both the magnitude and duration of thermal stress: a moderate anomaly
(e.g., 1°C) sustained for 12 weeks produces the same DHW value as a severe anomaly (e.g.,
3°C) lasting 4 weeks, though the biological impacts may differ. DHW estimates for each
indicative reef (see @GBRMPA2022) were then determined by taking the mean DHW value for all
pixels touching each reef outline.

::: {#tbl-gcms}
| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Results found in |
|-----------------------|-----------------------|-------------------------|-----------------------|
| EC-Earth3-Veg | 4.31 | High | **insert repo citation** |
| ACCESS-CM2 | 4.72 | High | **insert repo citation** |
| ACCESS-ESM1-5 | 3.87 | Intermediate | Main text |
| GFDL-CM4 | 2.9 | Low | **insert repo citation** |
| NorESM2-MM | 2.5 | Low | **insert repo citation** |

@@

Reef depth characteristics were extracted using 10m resolution bathymetry data from the
Great Barrier Reef Marine Park Authority (GBRMPA [-@GBRMPA2020]). For each reef, bathymetry
data (in raster format) were spatially matched with reef polygon features through geometric
intersection analysis. Depth values within each reef polygon were summarized using zonal
statistics to calculate minimum, mean, median, maximum, and standard deviation of depth
values. All depth values were converted to positive values (depths below sea level). The
median depth was used to provide an indicative depth value for each reef.

As bathymetry data are derived from satellite imagery they can be limited in their alignment
with the established GBRMPA reef outlines used in the investigation to characterise reef
locations. Quality control flags were assigned to identify potential data limitations: reefs
with no overlapping bathymetry data, reefs with minimum depth values above sea level
(indicating potential data artifacts), and reefs where bathymetry pixels covered less than
5% of the reef polygon area. These reefs with insufficient data coverage were removed from
analyses (842, or 22%, reefs of original 3806 reefs used in Bozec et al. [@bozec2025]).

**\[BG: I have updated now, does it make sense referring to the original 3806 reefs (I have referred to this 3806 set of reefs in the first paragraph of the methods)?\]**

While data adopted for the given assessment are said to be accurate up to 15m from Mean Sea
Level [@GBRMPA2020], there are many sources of uncertainty in the depth estimates used for
this study. There is potential misalignment between GBRMPA reef outlines, as well as the
usual challenges of deriving data from satellite imagery. In the case of bathymetry,
estimates are subject to different bottom types (rock, sand, seagrass, etc) and how these
affect reflectance, the water penetration depth of satellite sensors, influence of local
conditions (e.g., turbidity and wave activity), and many other factors. The resulting depth
estimates are therefore highly uncertain themselves and are only adopted as a useful
indication rather than an accurate and precise report of reef depths across the GBR.

@@

Current reef restoration and intervention-investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable restoration
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One of the main
constraints on deployment depth is visual identification of target sites from the surface
for deployment and ongoing monitoring, in combination with factors such as deployment method
and tourism return value. Our results suggest that of the assessed reefs ocurring within the
1-10m depth range (1736 reefs), 62% (1082) of reefs fall into the low and medium clusters
(across all GCMs at the GBRMPA management region scale), with many unable to maintain
positive carbonate budgets under future conditions. Reef intervention planning is a complex
and challenging decision making process [@anthony2020], and our results suggest that
deployment site depth is an important factor to consider when assessing the long-term
resilience of deployment sites.
